### 2018-04-11 Writing/disparity_review.tex
Suggestion for title
New reordered version of the introduction

\noindent{\Large \bf Disparity in morphological disparity analysis} 
% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
%TG: back to the drawing board. Maybe something with "best practices" as well.
\bigskip

@@

\section{Introduction}

Even classical scholars recognised that variation in the shape, form and functions of organisms often follows discrete patterns rather than a continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible life forms where each individual region are the result of a discrete combination of many variables.
For example, the mammalian carnivores design space can be seen as a combination of phylogenetic variables and their associated shared traits (i.e. the synapomorphies of the order Carnivora) as well as a combination of ecological, life history and behavioural variables (i.e. the ecological guild of meat eating mammals).

One commonly used approach to analyse these discrete patterns and to understand the processes that led to it is to analyse the disparity of the morphological charactersitics of the group, often referred to as morphological disparity analysis.
This type of analysis was popularised around the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] by studying the fossil record and focusing on broad questions such how does diversity evolve, are some body types more common that others or can shape evolve in all ``directions'' (of the multidimensional space).
Since then, an impressive amount of literature and the equally vast amount of methods have been used and the term ``disparity'' has become a close synonym of ``analysis of patterns of morphological diversity'' and has been used in a range of sub-fields in biology.
Disparity analysis has been used to tackle a vast array of questions in Evo-devo (Developmental constraints [CITE]; Relationship between form and function and evolution functional traits [CITE] or Scaling micro to macro [CITE];), in ecology or palaeoecology (intra-specific competition [CITE], niche occupants, speciation and competition[CITE] or macroecology (disparity in space)[CITE];)
or in macroevolution (response to extinction and selectivity [CITE], intra-specific competitive replacements [CITE] or more globally, tempo and mode of evolution [CITE]).

\subsection{What \textit{is} disparity?}

To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.

At the core of this debate a potential divide between whether disparity is a patterns or a processes: is there a difference between the disparity patterns (and how researchers characterise them) and how they were ultimately generated?
If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).

Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.
Thus, disparity methods or analysis can be seen as methods or analysis to study these patterns and eventually link them to macroevolutionary or macroecological processes

>>>>

\noindent{\Large \bf Disparity in morphological disparity analysis} 
% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
%TG: back to the drawing board. Maybe something with "best practices" as well.
% NC: how about something like "Analysing disparity in morphology data: best practice and future directions"
\bigskip

@@

% NC: OK I've chopped this up a lot as it felt like you were repeating yourself quite a bit (e.g. you said here is what kinds of analyses people did in the 90s, and what they do now, then you had a section on here is what we defined disparity as in the 1980s and here is what we do now - so I've combined it all) and could say the same thing in half the space. Feel free to revert back.
\section{Introduction}

% NC: New version
Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.

In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF).
Early disparity analyses focused on broad questions such as How does diversity evolve? Are some body types more common than others? Can shape evolve in all ``directions'' (of the multidimensional space)?
Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e. developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e. intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e. responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).

The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here). 
Here we use a broader definition and suggest that disparity is a pattern of variation.
Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.

% OLD VERSION
\subsection{old version of intro}
%Even classical scholars recognised that NC: I'm unsure we need this qualifier, just get right to it!
Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
% NC: I've tried to make the example clearer, but I'm unsure if it actually helps. Could cut the examples and go straight into the next sentence? Then the intro is just one paragraph.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as morphological disparity analysis.

This type of analysis was popularised in the 1980s and 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] through studies of the fossil record, focused on broad questions such as how does diversity evolve? Are some body types more common than others? Can shape evolve in all ``directions'' (of the multidimensional space)?

In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).

Since then, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity'' as a synonym for ``analysis of patterns of morphological diversity''.
Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e. developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e. intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e. responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).

What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).

Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).

Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.
Thus, disparity methods or analysis can be seen as methods or analysis to study these patterns and eventually link them to macroevolutionary or macroecological processes
